HumanCdGF_microbiota_predict_by_day

### 2016-09-09 code/HumanCdGF_microbiota_predict.Rmd

#random forest regression model to determine Day 0 OTU correlated with increased CFU
seed <- 18
n_trees <- 2000
set.seed(seed)

@@

##Can we predict moribundity with the community structure at Day 0?
#### randomForest
```{r fig.width=10, fig.height=5}
meta_file$Euth_Early <- factor(ifelse(meta_file[ , 'Early_Euth'] == FALSE, 1, 0))

@@

rf_early_euth <- randomForest(Euth_Early ~ ., data = Predict_early_euth_df, 
                              importance = TRUE, ntree = n_trees)
rf_euth_import <- data.frame(importance = head(sort(rf_early_euth$importance[ , 'MeanDecreaseAccuracy'], decreasing = T), 10))

rf_early_euth_top_otu <- randomForest(Euth_Early ~ ., 
                                      data = Predict_early_euth_df[, c(rownames(rf_euth_import), 'Euth_Early')], 
                              importance = TRUE, ntree = n_trees)

unpredicted_top_otu <- Predict_early_euth_df_names[Predict_early_euth_df$Euth_Early != 
                                               rf_early_euth_top_otu$predicted]
rf_euth_import_top_otu <- data.frame(importance = sort(rf_early_euth_top_otu$importance[ , 'MeanDecreaseAccuracy'], 
                                                       decreasing = T))
ROC_euth_top_otu <- roc(as.numeric(Predict_early_euth_df$Euth_Early), as.numeric(rf_early_euth_top_otu$predicted))
roc_plot_top_otu <- data.frame(Specificity = 1 - ROC_euth_top_otu$specificities, 
                       Sensitivity = ROC_euth_top_otu$sensitivities)

euth_rf_plot_top_otu <- ggplot(data = roc_plot_top_otu, aes(x = Specificity, y = Sensitivity)) + 
                      geom_path() + geom_abline(intercept = 0, slope = 1, 
                                                linetype = 'dashed', alpha = 0.5)
euth_import_plot_top_otu <- ggplot(data = rf_euth_import_top_otu, 
                           aes(x = factor(rownames(rf_euth_import_top_otu)), y = importance)) + 
        geom_point() + scale_x_discrete(limits = rev(rownames(rf_euth_import_top_otu))) +
        labs(x= '', y = '% Increase in MSE') + theme_bw() + coord_flip()
# title = 'OTU Importance in Prediction of Day Euthanized from Day 0 Community',

ggdraw()+
  draw_plot(euth_rf_plot_top_otu, 0, 0, 0.5, 1) + 
  draw_plot(euth_import_plot_top_otu, 0.5, 0, 0.5, 1)
```

Confusion Matrix  
`r kable(rf_early_euth_top_otu$confusion)`

OOB error rate = `r 100*tail(rf_early_euth_top_otu$err.rate[ , 1], 1)`

Mice which were euthanized early, but the model predicts the will were not (Mouse Tag - Cage):
`r unpredicted_top_otu`

#### AUCRF

```{r fig.width=6, fig.height=5}

>>>>

#random forest regression model to determine Day 0 OTU correlated with increased CFU
seed <- 18
n_trees <- 2001
set.seed(seed)

@@

##Can we predict moribundity with the community structure at Day 0?

```{r fig.width=6, fig.height=5}
meta_file$Euth_Early <- factor(ifelse(meta_file[ , 'Early_Euth'] == FALSE, 1, 0))

@@

### 2016-09-09 code/HumanCdGF_microbiota_predict_by_day.Rmd

# read in files
meta_file   <- read.table('~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt', sep = '\t',
                        header = T)
shared_file <- read.table('~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared',
                          sep='\t',header = T, row.names = 'Group')
tax_file <- read.table('~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_cdiff.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.cons.taxonomy',
                       sep='\t',header = T, row.names = 'OTU')
# function to label otu with tax
source('~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/tax_level.R')

# identify c difficile otus
cdiff_otu <- rownames(tax_file[grep('Clostridium_XI', tax_file$Taxonomy),])
# remove samples not used or with other cdiff strains
meta_file <- 
  meta_file %>% 
  filter(sample_type == 'stool', cdiff_strain == 431) %>% 
  select(Mouse_ID, group, day, cdiff_cfu, Euthanized) %>% 
  mutate(log_cfu = log10(cdiff_cfu + 1)) %>% 
  filter(!is.na(log_cfu))

# create shared with relative abundances over 1%
mean_otu_counts <- unique(apply(shared_file, 1, sum))
rel_abund_shared <- 100 * shared_file / mean_otu_counts
rel_abund_shared_01 <- rel_abund_shared[ , apply(rel_abund_shared, 2, max) > 1]
rel_abund_shared_01$group <- rownames(rel_abund_shared_01)

# create df with day 0 communities with column of mouse_id
day_0_otu_by_mouse <- 
  meta_file %>% 
  filter(day == 0 ) %>% 
  select(group, Mouse_ID) %>% 
  left_join(rel_abund_shared_01) %>% 
  select(-group)
# create df with cfu merge to shared
cfu_by_day_comm <-
  meta_file %>% 
  filter(day %in% 1:10) %>% 
  select(group, day, log_cfu) %>% 
  left_join(rel_abund_shared_01) %>% 
  select(-group, -one_of(cdiff_otu))

@@

RandomForest_by_day <- function(i, community_day_0){
  if(!is.logical(community_day_0)){
    stop('Input for community  - Need community_day_0 = T/F')
  } else {
    rf_data <- if(community_day_0 == TRUE){
      meta_file %>%
        filter(day == 1) %>% 
        select(log_cfu, Mouse_ID) %>%
        inner_join(day_0_otu_by_mouse) %>% 
        select(-Mouse_ID)
    } else {
      cfu_by_day_comm %>% 
        filter(day == 1) %>% 
        select(-day)
    }

@@

```{r fig.width=10, fig.height=10}
meta_file %>% 
  select(day, log_cfu, Mouse_ID) %>% 
  full_join(day_0_otu_by_mouse) %>% 
  select(day, log_cfu, one_of(predictive_otus)) %>% 
  filter(!day == 0) %>% 
  gather('otu', 'abundance', contains('Otu')) %>% 
  filter(!is.na(abundance)) %>% 
  full_join(predictive_taxa) %>% 
  ggplot(aes(x = abundance, y = log_cfu, color = factor(day))) + geom_point() + 
    facet_wrap(~ tax_label, scales = 'free_x') +
    labs( x = 'Relative Abundance', y = 'CFU (log10)')

@@

```{r fig.width=10, fig.height=10}
meta_file %>% 
  select(day, log_cfu, Mouse_ID) %>% 
  full_join(day_0_otu_by_mouse) %>% 
  select(day, log_cfu, one_of(predictive_otus)) %>% 
  filter(!day == 0) %>% 
  gather('otu', 'abundance', contains('Otu')) %>% 
  filter(!is.na(abundance)) %>% 
  full_join(predictive_taxa) %>% 
  ggplot(aes(x = log10(abundance), y = log_cfu, color = factor(day))) + geom_point() + 
    facet_wrap(~ tax_label, scales = 'free_x') +
    labs( x = 'Relative Abundance (log10)', y = 'CFU (log10)')

@@

corr_df_day_0_comm <- meta_file %>% 
  select(day, log_cfu, Mouse_ID) %>% 
  full_join(day_0_otu_by_mouse) %>% 
  filter(!day == 0) %>% 
  select(-day, -Mouse_ID)

@@

>>>>

# read in files
meta_file   <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt'
shared_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
tax_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_all.an.cons.taxonomy'
tax_function <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/tax_level.R'

# read in files
meta_file   <- read.table(meta_file, sep = '\t', header = T)
shared_file <- read.table(shared_file, sep = '\t', header = T, row.names = 'sample_id')
tax_file <- read.table(tax_file, sep = '\t', header = T, row.names = 'OTU')
source(tax_function)

# identify c difficile otus
cdiff_otu <- rownames(tax_file[grep('Clostridium_XI', tax_file$Taxonomy),])
# remove samples not used or with other cdiff strains
meta_file <- 
  meta_file %>% 
  filter(cdiff_strain == 431) %>% 
  select(mouse_id, cage_id, sample_id, day, cdiff_cfu, Early_Euth, log_cfu) %>% 
  filter(!is.na(log_cfu))

# create shared with relative abundances over 1%
mean_otu_counts <- unique(apply(shared_file, 1, sum))
rel_abund_shared <- 100 * shared_file / mean_otu_counts
rel_abund_shared_01 <- rel_abund_shared[ , apply(rel_abund_shared, 2, max) > 1]
rel_abund_shared_01$sample_id <- rownames(rel_abund_shared_01)

# create df with day 0 communities with column of mouse_id
day_0_otu_by_mouse <- 
  meta_file %>% 
  filter(day == 0 ) %>% 
  select(sample_id, mouse_id) %>% 
  left_join(rel_abund_shared_01) %>% 
  select(-sample_id)
# create df with cfu merge to shared
cfu_by_day_comm <-
  meta_file %>% 
  filter(day %in% 1:10) %>% 
  select(sample_id, day, log_cfu) %>% 
  left_join(rel_abund_shared_01) %>% 
  select(-sample_id, -one_of(cdiff_otu))

@@

RandomForest_by_day <- function(i, community_day_0){
  if(!is.logical(community_day_0)){
    stop('Input for community  - Need community_day_0 = T/F')
  } else {
    rf_data <- if(community_day_0 == TRUE){
      meta_file %>%
        filter(day == 1) %>% 
        select(log_cfu, mouse_id) %>%
        inner_join(day_0_otu_by_mouse) %>% 
        select(-mouse_id)
    } else {
      cfu_by_day_comm %>% 
        filter(day == 1) %>% 
        select(-day)
    }

@@

```{r fig.width=10, fig.height=10}
meta_file %>% 
  select(day, log_cfu, mouse_id) %>% 
  full_join(day_0_otu_by_mouse) %>% 
  select(day, log_cfu, one_of(predictive_otus)) %>% 
  filter(!day == 0) %>% 
  gather('otu', 'abundance', contains('Otu')) %>% 
  filter(!is.na(abundance)) %>% 
  full_join(predictive_taxa) %>% 
  ggplot(aes(x = abundance, y = log_cfu, color = factor(day))) + geom_point() + 
    facet_wrap(~ tax_label, scales = 'free_x') +
    labs( x = 'Relative Abundance', y = 'CFU (log10)')

@@

```{r fig.width=10, fig.height=10}
meta_file %>% 
  select(day, log_cfu, mouse_id) %>% 
  full_join(day_0_otu_by_mouse) %>% 
  select(day, log_cfu, one_of(predictive_otus)) %>% 
  filter(!day == 0) %>% 
  gather('otu', 'abundance', contains('Otu')) %>% 
  filter(!is.na(abundance)) %>% 
  full_join(predictive_taxa) %>% 
  ggplot(aes(x = log10(abundance), y = log_cfu, color = factor(day))) + geom_point() + 
    facet_wrap(~ tax_label, scales = 'free_x') +
    labs( x = 'Relative Abundance (log10)', y = 'CFU (log10)')

@@

corr_df_day_0_comm <- meta_file %>% 
  select(day, log_cfu, mouse_id) %>% 
  full_join(day_0_otu_by_mouse) %>% 
  filter(!day == 0) %>% 
  select(-day, -mouse_id)

@@

```{r}
# function to compare rel abund of otu with cdiff cfu
plot_otu_w_cfu <- function(OTU){
  rel_abund_otu <- data.frame(rel_abund = rel_abund_shared_01[,OTU],
                              sample_id = rownames(rel_abund_shared_01))
  plot <- meta_file %>% 
  select(sample_id, day, cage_id, log_cfu) %>% 
  left_join(rel_abund_otu) %>% 
  select(cage_id, day, log_cfu, rel_abund) %>% 
  filter(day %in% 0:10) %>%
  group_by(cage_id, day) %>% 
  summarise(abund = mean(rel_abund), cfu = mean(log_cfu)) %>% 
  gather(data_type, values, abund:cfu) %>% 
  ggplot(aes(x = factor(day), y = values, group = data_type, color = data_type)) +
            facet_wrap(~cage_id) + 
            geom_line() + scale_y_log10()
}
